Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-02-11 multi_body_interactions/papers/theory/reviews_NJP/referee_1_response.tex

\begin{enumerate}
\item Concerning:

  \blue{The authors study the effective on-site interactions for the
    description of several fermionic particles confined in a very deep
    optical lattice. The focus is on alkaline earth atoms, where the
    s-wave interaction strength is independent on the nuclear spin,
    and on all atoms in the lowest state of the trapping
    potential. The manuscript is an extended version on the
    theoretical tools applied in the comparison between experiment and
    theory of Ref. [34].  The analysis is very important and
    timely. However, in the present form, the manuscript has some
    significant short comings, which have to be improved before a
    decision about publication can be made.}

  \blue{One of the main points of criticism is that the manuscript is
    very poorly written and very difficult to access. A few examples
    are listed below:}

  We thank the referee for the succinct summary of our work, and for
  recognizing that our analysis is ``very important and timely''.  We
  hope that thew revisions outlined below have made our manuscript
  more accessible, and that we have addressed the shortcomings
  identified by the referee to a satisfactory degree.

@@

\item Concerning the generality of our multi-body theory:

  \blue{(i) The authors try to be rather general in their description
    such as title, abstract and introduction. During the description
    of the results, they start to apply several approximations, which
    are neither mentioned before or justified.  As an example: the
    authors claim to derive a general low energy theory for orbital
    and nuclear spin stats. However, in Eq.(10) the authors suddenly
    restrict the analysis to a low density of excitations in the ``e''
    orbital state. This restriction is never explained nor motivated.
    In addition, the restriction to a single occupation of the nuclear
    spin state is never motivated or justified but just introduced at
    some point. The referee understands, that these restrictions are
    justified for the comparison with the experiment in
    Ref.[34]. However, it should be clearly pointed, that the analysis
    is not general and adapted to the description to these experiments
    in the abstract and introduction. Furthermore, a short description
    of the experiment would be mandatory for an independent
    manuscript.}

  We thank the referee for pointing out the possibility to mislead
  readers on the generality of our theory.  The referee is correct
  that we do not ``derive a general low energy theory for orbital and
  nuclear spin states'', and we agree on the importance of making the
  scope of our technical contributions clear.  In particular, our
  theory does not consider multiple occupation of a nuclear spin state
  on any lattice site.  To facilitate comparison with experiment, we
  additionally consider a restriction of our theory to the space of
  states with at most one orbital excitation per lattice site, as
  e.g.~in Eq.~(10) (page 8) and section IV (page 18).  The full theory
  developed in section III (page 8), however, is completely general
  with respect to the number of orbital excitations per lattice site.
  We also thank the referee for pointing out that a short description
  of relevant experimental procedures would make our manuscript more
  self-contained, and we recognize the value in doing so.

@@

  To clarify the limitations of our theory, we have added a sentence
  to our abstract, such that the original text (page 1, line 34):

  \red{... nuclear spin levels.  We fully characterize...}

  now reads:

  \green{... nuclear spin levels.  Our theory is limited to the
    experimental regime of (i) a deep lattice, with (ii) at most one
    atom occupying each nuclear spin state on any lattice site.  The
    latter restriction is a consequence of initial ground-state
    preparation.  We fully characterize...}

  We now explicitly mention the ground-state preparation in the
  introduction, replacing (page 4, line 22):

@@

  and we have appended an additional sentence to the same paragraph,
  previously ending with the text (page 4, line 31):

  \red{... through the exquisite capabilities with OLCs.}

  and now:

  \green{... through the exquisite capabilities with OLCs.  Our theory
    is limited to the experimental regime of at most one atom
    occupying each nuclear spin state on any lattice site, which is a
    consequence of the experimental protocol which starts with all
    atoms in the ground state.}

@@

  In order to motivate a later restriction to one orbital excitation
  per lattice site, we have rearranged and rewritten some of the next
  paragraph in the introduction, changing the text (page 4, line 33):

  \red{Though effective multi-body interactions have previously been
    studied in the context of harmonically [35, 36] and
    lattice-confined [37] neutral bosons prepared in a single
    hyperfine state, our work deals for the first time with fermions
    that have internal degrees of freedom and multiple collisional
    parameters.  Nonetheless, due to the SU($N$) symmetry of these
    collisions, we are able to find a simple way to express effective
    multi-body Hamiltonians and fully characterize their low-lying
    many-body energy eigenstates.  While some past work has detected
    experimental signatures of multi-body interactions in the form of
    quantum phase revivals [38], we perform a direct comparison of the
    many-body interaction energies predicted by our low-energy
    effective theory with experimental measurements of
    density-dependent atomic energy spectra performed in ref.~[34],
    similarly to the measurements with bosons performed in ref.~[39].}

  which now reads:

  \green{Though effective multi-body interactions have previously been
    studied in the context of harmonically [35, 36] and
    lattice-confined [37] neutral bosons prepared in a single
    hyperfine state, our work deals for the first time with fermions
    that have internal degrees of freedom and multiple collisional
    parameters.  Some past work has detected experimental signatures
    of multi-body interactions in the form of quantum phase revivals
    [39].  We instead compare the many-body interaction energies
    predicted by our low-energy effective theory to the experimental
    measurements of the density-dependent orbital excitation spectra
    performed in ref.~[34], similarly to the measurements with bosons
    performed in ref.~[38].  To facilitate this comparison of
    excitation spectra and to characterize the low-lying excitations
    in our effective theory, we consider a restriction of our theory
    to states with at most one orbital excitation per lattice site.
    In this case, we find that the SU($N$) symmetry of atomic
    collisions allow the effective multi-body interactions to take a
    remarkably simple form.}

  We also clarify the limitations of our theory in the text between
  Eqs.~(8) and (9), as discussed later in point \ref{pt:confusion},
  and have made our technical contributions more explicit in the
  description of our main results (in section II), replacing (page 7,
  line 48):

@@

  \red{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align*}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} H_M, \tag{9}
    \end{align*}
    where $H_M$ is an $M$-body Hamiltonian, $I$ is the total nuclear
    spin of each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum
    terminates at $2I+1$ because this is the largest number of atoms
    which may initially occupy a single lattice site.  Focusing on the
    low-lying orbital state excitations of this theory, we find that
    SU($N$) symmetry allows us to express multi-body Hamiltonians in
    the simple form}

  by:

  \green{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align*}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} \sum_{p\ge1} H_M^{(p)},
      \tag{9}
    \end{align*}
    where $H_M^{(p)}$ is an $M$-body Hamiltonian of order $p$ in the
    coupling constants $G_X$, and $I$ is the total nuclear spin of
    each atom (e.g.~$I=9/2$ for ${}^{87}$Sr).  The sum terminates at
    $2I+1$ because this is the largest number of atoms which may
    initially occupy a single lattice site.  We explicitly compute all
    of the $M$-body Hamiltonians $H_M\equiv\sum_p H_M^{(p)}$ through
    order $p=3$, yielding effective two-, three-, and four-body
    interactions.  To (i) facilitate a comparison with the
    experimental measurements of many-body orbital excitation spectra
    performed in ref.~[34] and (ii) characterize the low-lying
    excitations in our effective theory, we also consider a
    restriction of the multi-body Hamiltonians $H_M$ to states with at
    most one orbital excitation per lattice site.  Under this
    restriction, we find that the SU($N$) symmetry of atomic
    collisions allows us to express all multi-body Hamiltonians in the
    simple form}

@@

  Finally, we have modified some text in our summary to emphasize that
  our main theory is general with respect to orbital excitations, and
  that the restriction to one orbital excitation per lattice site is
  made to facilitate comparison with experiments; specifically, we
  have changed (page 28, line 40):

  \red{Working in the deep-lattice limit, we have derived a low-energy
    effective theory of such atoms.  This theory exhibits emergent
    multi-body interactions that inherit the SU($N$) symmetry of the
    atoms' bare, two-body interactions.  The SU($N$) symmetry of
    $M$-body Hamiltonians in the effective theory has allowed us to
    express them in a simple form, and fully characterize their
    corresponding eigenstates and spectra.}

  to:

  \green{Working in the deep-lattice limit and the experimental regime
    of at most one atom occupying each nuclear spin state on any
    lattice site, we have derived a low-energy effective theory of
    these atoms.  Our theory exhibits emergent multi-body interactions
    that inherit the SU($N$) symmetry of the bare two-body
    interactions.  Considering a restriction of our theory to the
    subspace of at most one orbital excitation per lattice site, we
    found that the SU($N$) symmetry of all $M$-body Hamiltonians
    allowed us to express them in a simple form, and to fully
    characterize their eigenstates and spectra.}

  We hope that these changes clarify the scope of our work, in
  particular highlighting that (i) our theory is limited to the case
  of atoms in distinct nuclear spin states on each lattice site, and
  (ii) the theory in section III applies to any number of orbital
  excitations, despite the restriction to the single-excitation case
  for the sake of comparison with experiments in section IV.

@@

  Concerning a description of the experiment in ref.~[34], we have
  added a summary of the relevant experimental procedures to the
  beginning of section II, such that the previous opening (page 5,
  line 19):

  \red{Consider a collection of fermionic atoms with total nuclear
    spin $I$ and two atomic orbital states loaded into a
    translationally invariant 3-D optical lattice.  While an external
    trapping potential will generally break discrete translational
    symmetry of the lattice, any background inhomogeneity can be made
    negligible by spectroscopically addressing a sufficiently small
    region of the lattice [34].  Throughout this paper, we work
    strictly in the deep-lattice regime with negligible tunneling
    between lattice sites.  We also assume a so-called
    ``magic-wavelength'' lattice for which both atomic orbital states
    experience identical lattice potentials [40].}

  now reads:

  \green{The work in this paper is closely tied to the experimental
    work reported in ref.~[34]; we begin with a short summary of the
    relevant experimental procedures therein.  The experiment begins
    by preparing a degenerate gas of $10^4$-$10^5$ (fermionic)
    ${}^{87}$Sr atoms in a uniform mixture of their ten nuclear spin
    states and at $\sim0.1$ of their fermi temperature ($\sim10$
    nanokelvin) [8, 40].  This gas is loaded into a primitive cubic
    optical lattice at the ``magic wavelength'' for which both ground
    (${}^1S_0$) and first-excited (${}^3P_0$) electronic orbital
    states of the atoms experience the same lattice potential [41].
    Lattice depths along the principal axes of the lattice are roughly
    equal in magnitude, with a geometric mean that can be varied from
    30 to 80 $E_{\t{R}}$, where
    $E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon
    recoil energy of the atoms (with the reduced Planck constant
    $\hbar=1$ throughout this paper).  These lattice depths are
    sufficiently large as to neglect tunneling on the time scales
    relevant to the experiment.  The temperature of the atoms is also
    low enough to neglect thermal occupation of motional states
    outside the ground-state manifold.}

  \green{Once loaded into an optical lattice, atoms are addressed by
    an external (``clock'') interrogation laser with an ultranarrow
    (26 mHz) linewidth, detuned by $\Delta$ from the single-atom
    ${}^1S_0-{}^3P_0$ transition frequency $\omega_0$.  After a fixed
    interrogation time, the experiment turns off the interrogation
    laser, removes all ground-state (${}^1S_0$) atoms from the
    lattice, and uses absorption imaging to count the remaining
    excited-state (${}^3P_0$) atoms.  Non-interacting atoms in
    singly-occupied lattice sites feature the typical single particle
    lineshape peaked at $\Delta=0$.  The lineshapes of
    multiply-occupied lattice sites, meanwhile, are shifted by
    inter-atomic interactions, which results in spectroscopic peaks
    (i.e.~local maxima in excited-state atom counts) away from
    $\Delta=0$.  A sweep across different detunings $\Delta$ (on the
    scale of inter-atomic interaction energies) thus constitutes a
    measurement of the many-body orbital excitation spectrum.  We note
    that this spectroscopic protocol addresses only singly-excited
    orbital states of lattice sites.  Doubly-excited states are
    manifestly off-resonant for single-photon absorption processes as
    $\abs{\Delta}\ll\omega_0$, while multi-photon absorption processes
    are always off-resonant due to (i) the interaction-induced
    non-linearity ($\sim$kHz) of the orbital excitation energies, and
    (ii) the ultranarrow linewidth ($\sim$mHz) of the interrogation
    laser.}

  \green{While an external trapping potential will generally break
    discrete translational symmetry of the lattice, any background
    inhomogeneity can be made negligible by spectroscopically
    addressing a sufficiently small region of the lattice [34].
    Throughout this paper, we work strictly in the deep-lattice regime
    with negligible tunneling between lattice sites.  We also neglect
    any lattice inhomogeneities and assume that both atomic orbital
    states (i.e.~${}^1S_0$ and ${}^3P_0$) experience identical lattice
    potentials.}

  To reflect this change made to section II, we have modified the
  description of section II on page 4, line 52, previously:

@@

  We thank the referee for their careful reading of our manuscript,
  and for pointing out that our discussion in the paragraph between
  Eqs.~(8) and (9) is poorly worded.  The paragraph in question
  appears on page 7, lines 20--47:

  \red{We are interested in the ground and low-lying orbital
    eigenstates of several interacting atoms on a single lattice site,
    as such states can be readily prepared and coherently addressed in
    current experiments with ultracold atoms [34].  Although these
    experiments operate at temperatures sufficiently low for
    negligible thermal occupation of excited motional states, simply
    neglecting excited single-particle motional states fails to
    reproduce the observed spectra of multiply-occupied lattice sites.
    The reason for this failure lies in the fact that the bare
    two-body Hamiltonian $H_{\t{int}}$ in (7) is not diagonal in the
    basis of single-particle motional states.  Even at zero
    temperature, therefore, the presence of excited motional states
    modifies many-body atomic energy spectra when the magnitude of
    off-diagonal elements of $H_{\t{int}}$ approach the scale of
    low-lying single-particle motional state energy gaps.  To address
    this problem, we develop a low-energy effective theory for
    ultracold two-level fermionic atoms on a lattice.  As our work is
    motivated by progress in experiments which initialize all atoms in
    identical (i.e.~ground) orbital and motional states, to simplify
    our theory we assume that any $N$ atoms on a single lattice site
    occupy $N$ nuclear spin states.  Multiple occupation of a single
    nuclear spin state on the same lattice site is initially forbidden
    by fermionic statistics, and subsequently cannot be violated in
    the absence of tunneling and hyperfine coupling.}

  We have re-written this paragraph in its entirety to walk through
  the arguments and explanations therein more carefully, replacing it
  with the following text:

  \green{Current experiments with ${}^{87}$Sr can prepare up to five
    atoms in the same (ground) orbital state on a single lattice site,
    and coherently address states with a single orbital excitation per
    lattice site [34].  At ultracold temperatures well below the
    single-atom motional excitation energies $\Delta_n\equiv E_n-E_0$
    for $n>0$, atoms only occupy their motional ground state in the
    lattice.  For this reason, it is common to map the description of
    these atoms onto a single-band Hubbard model that captures all
    dynamics within the subspace of motional ground states of
    non-interacting atoms, i.e.~with single-atom wavefunctions
    $\phi_{i,0}$.  Interactions, however, modify atoms' motional
    ground-state wavefuctions.  The true motional ground state of a
    collection of interacting atoms is then an admixture of the
    non-interacting motional eigenstates, and a naive Hubbard model
    that assumes atomic wavefunctions $\phi_{i,0}$ will fail to
    reproduce the interacting atoms' orbital excitation spectrum.
    Formally, corrections to the spectrum of interacting atoms can be
    accounted for by a perturbative treatment of far-off-resonant
    terms in the interaction Hamiltonian of (7) that create atoms in
    excited motional states,
    e.g.~$\sim\c_{n\mu s}^\dag \c_{0,\nu t}^\dag \c_{0,\nu r}
    \c_{0,\mu q}$ with $n>0$.  These corrections can be understood
    through interaction-induced {\it virtual} occupation of higher
    bands (i.e.~excited motional states), which becomes relevant as
    more atoms occupy the same lattice site, such that their
    interaction energy becomes non-negligible compared to the motional
    excitation energies $\Delta_n$.}

  \green{In order to recast interaction-induced modifications to
    orbital excitation spectra as corrections to the simple Hubbard
    model (i.e.~computed using the non-interacting ground-state
    wavefunctions $\phi_{i,0}$), we develop a low-energy effective
    theory of interacting AEAs in a deep lattice.  To simplify our
    theory, we assume that any $N$ atoms on a single lattice site
    occupy distinct nuclear spin states.  This assumption applies for
    any experimental protocol in which all atoms are initially
    prepared in their orbital and motional ground states (as e.g.~in
    ref.~[34]).  In this case, multiple occupation of a single nuclear
    spin state on any given lattice site is initially forbidden by
    fermionic statistics.  Subsequent violation of this condition
    cannot occur in the absence of inter-site effects or hyperfine
    coupling between nuclear spin states, as is the case in the
    experiment in ref.~[34].}

@@

  As the referee points out, Eqs.~(10) and (42) in the original
  version of our manuscript are identical.  In order to avoid
  redundancy while still providing a preview of our main results, we
  have replaced the following text in section II (page 8, line 10):

### 2019-02-11 multi_body_interactions/papers/theory/paper.tex
merged some changes with AMR's suggestions

\usepackage{setspace} % to modify spacing in table of contents

@@

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  Motivated in
  particular by recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent multi-body SU($N$)-symmetric
  interactions, where $N$ is the number of atomic nuclear spin levels.
  Our theory is limited to the experimental regime of
  \begin{enumerate*}
  \item a deep lattice, with
  \item at most one atom occupying each nuclear spin state on any
    lattice site,
  \end{enumerate*}
  where the latter restriction is a consequence of initial
  ground-state preparation.  We fully characterize the low-lying
  excitations in our effective theory, and compare predictions of
  many-body interaction energies with direct measurements of many-body
  excitation spectra in an optical lattice clock.  Our work makes the
  first step in enabling a controlled, bottom-up experimental
  investigation of multi-body SU($N$) physics.
\end{abstract}

@@

\begin{spacing}{1.3}
  \tableofcontents
\end{spacing}

@@

In this work, we investigate the first experimental capabilities with
ultracold fermionic AEAs to prepare high-density samples in a 3-D
lattice with multiple occupation of individual lattice sites
\cite{goban2018emergence}.  Specifically, we consider ground-state
preparation of isolated few-body systems in the deep-lattice limit,
and carry out a bottom-up investigation of emergent multi-body
interactions on multiply-occupied lattice sites.  These multi-body
interactions appear in a low-energy effective theory of the atoms, and
inherit the SU($N$) symmetry of their bare, pair-wise interactions,
thereby enabling experimental studies of multi-body SU($N$) physics
through the exquisite capabilities with OLCs.  Our theory is limited
to the experimental regime of at most one atom occupying each nuclear
spin state on any lattice site, which is a consequence initial
ground-state preparation.

Though effective multi-body interactions have previously been studied
in the context of harmonically \cite{johnson2012effective,
  yin2014universal} and lattice-confined \cite{johnson2009effective}
neutral bosons prepared in a single hyperfine state, our work deals
for the first time with fermions that have internal degrees of freedom
and multiple collisional parameters.  We perform a direct comparison
between the many-body interaction energies predicted by our low-energy
effective theory and the experimental measurements of the
density-dependent orbital excitation spectra performed in
ref.~\cite{goban2018emergence}, similarly to the measurements with
bosons performed in ref.~\cite{mark2011precision}.  To facilitate this
comparison of excitation spectra and characterize the low-lying
excitations in our effective theory, we consider a restriction of the
theory to states with at most one orbital excitation per lattice site,
and find that the SU($N$) symmetry of atomic collisions allow
effective multi-body interactions to take a remarkably simple form.
We note that past work has also detected experimental signatures of
multi-body interactions in the form of quantum phase revivals
\cite{will2010timeresolved}.

@@

The work in this paper is closely tied to the experimental work in
ref.~\cite{goban2018emergence}; we begin with a short summary of the
relevant experimental procedures therein.  The experiment begins by
preparing a degenerate gas of $10^4$-$10^5$ (fermionic) ${}^{87}$Sr
atoms in a uniform mixture of their ten nuclear spin states and at
$\sim0.1$ of their fermi temperature ($\sim10$ nanokelvin)
\cite{campbell2017fermidegenerate, marti2018imaging}.  This gas is
loaded into a primitive cubic optical lattice at the ``magic
wavelength'' for which both ground (${}^1S_0$) and first-excited
(${}^3P_0$) electronic orbital states of the atoms experience the same
lattice potential \cite{ye2008quantum}.  Lattice depths along the
principal axes of the lattice are roughly equal in magnitude, with a
geometric mean that can be varied from 30 to 80 $E_{\t{R}}$, where
$E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon recoil
energy of the atoms (with the reduced Planck constant $\hbar=1$
throughout this paper).  These lattice depths are sufficiently large
as to neglect tunneling on the time scales relevant to the experiment.
The temperature of the atoms is also low enough to neglect thermal
occupation of states outside the ground-state manifold.

Once loaded into an optical lattice lattice, atoms are addressed by an
external (``clock'') interrogation laser with an ultranarrow (26 mHz)
linewidth, detuned by $\Delta$ from the single-atom ${}^1S_0-{}^3P_0$
transition frequency $\omega_0$.  After a fixed interrogation time,
the experiment turns off the interrogation laser, removes all
ground-state (${}^1S_0$) atoms from the lattice, and uses absorption
imaging to count any remaining excited-state (${}^3P_0$) atoms.
Non-interacting atoms in singly-occupied lattice sites undergo Rabi
oscillations when $\Delta=0$ during the interrogation time.  The
orbital excitation spectrum of multiply-occupied lattice sites,
however, is shifted by inter-atomic interactions, which results in
spectroscopic peaks (i.e.~local maxima in excited-state atom counts)
away from $\Delta=0$.  A sweep across different detunings $\Delta$ (on
the scale of inter-atomic interaction energies) thus constitutes a
measurement of the many-body orbital excitation spectrum.  We note
that this spectroscopic protocol addresses only singly-excited orbital
states of any lattice site.  Doubly-excited states are manifestly
off-resonant for single-photon absorption processes as
$\abs{\Delta}\ll\omega_0$, while multi-photon absorption processes are
always off-resonant due to
\begin{enumerate*}
\item the non-linearity ($\sim$kHz) of orbital excitation energies for
  a collection of interacting atoms, and
\item the ultranarrow linewidth ($\sim$mHz) of the interrogation
  laser.
\end{enumerate*}

Consider now a collection of fermionic atoms with total nuclear spin
$I$ and two atomic orbital states loaded into a translationally
invariant 3-D optical lattice.  While an external trapping potential
will generally break discrete translational symmetry of the lattice,
any background inhomogeneity can be made negligible by
spectroscopically addressing a sufficiently small region of the
lattice \cite{goban2018emergence}.  Throughout this paper, we work
strictly in the deep-lattice regime with negligible tunneling between
lattice sites, and assume that both atomic orbital states experience
identical lattice potentials.  The single-particle Hamiltonian of the
atoms can then be written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,-I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom in
motional state $n$.  In a harmonic trap approximation we would have
$E_n=\p{3/2+n_\x+n_\y+n_\z}\omega$ for an on-site angular trap
frequency $\omega$, but in general the aharmonicity of the lattice
potential will cause a non-negligible shift in motional state
energies.

@@

Second, \eqref{eq:H_int_fields} is only the first term in a low-energy
expansion of two-body interactions in effective field theory, which
generally includes additional terms containing derivatives of field
operators.  Derivative terms correspond to the dependence of two-body
scattering on the relative momentum $k$ of particles involved, with
$k\to0$ in the zero-energy limit.  In the present case of $s$-wave
scattering, the leading dependence of the two-body interaction
Hamiltonian on the relative momentum $k$ can be captured by use of an
energy-dependent pseudo-potential, which amounts to using a
$k$-dependent effective scattering length \cite{blume2002fermi}.  This
effective scattering length can be determined by expanding the
$s$-wave collisional phase shift in powers of the relative momentum
$k$ \cite{giorgini2008theory, flambaum1999analytical}.  Details of
this expansion will depend on the characteristic length scale of
finite-range interactions.  In our work, these corrections to
\eqref{eq:H_int_fields} will be relevant only for the calculation of
two-body interaction energies, appearing at third order in the
coupling constants $G_X$.  As we are primarily interested in $M$-body
interactions for $M\ge3$, we defer this calculation to Appendix
\ref{sec:momentum_dependence}.  We note that our approach of using an
unregularized contact potential, renormalizing coupling constants, and
separately accounting for momentum-dependent scattering is essentially
the same as that used for similar calculations in
refs.~\cite{johnson2012effective, yin2014universal,
  johnson2009effective}.  While this approach does not provide insight
into the microscopic structure of inter-atomic interactions, it is
suitable for the phenomenological description of these interactions,
and particular for our eventual development of a low-energy effective
theory.

@@

(and similarly $K^\ell_m$) becomes sharply peaked at $\ell\approx m$
as $\ell$ and $m$ get large.  The asymptotic behavior of
$\alpha_3^{(3)}$ at large $\ell,m,n$ is therefore

>>>>

@@

\begin{abstract}
  Rapid advancements in the experimental capabilities with ultracold
  alkaline-earth-like atoms (AEAs) bring to a surprisingly near term
  the prospect of performing quantum simulations of spin models and
  lattice field theories exhibiting SU($N$) symmetry.  Motivated in
  particular by recent experiments preparing high density samples of
  strongly interacting ${}^{87}$Sr atoms in a three-dimensional
  optical lattice, we develop a low-energy effective theory of
  fermionic AEAs which exhibits emergent multi-body SU($N$)-symmetric
  interactions, where $N$ is the number of atomic nuclear spin levels.
  Our theory is limited to the experimental regime of
  \begin{enumerate*}
  \item a deep lattice, with
  \item at most one atom occupying each nuclear spin state on any
    lattice site.
  \end{enumerate*}
  The latter restriction is a consequence of initial ground-state
  preparation.  We fully characterize the low-lying excitations in our
  effective theory, and compare predictions of many-body interaction
  energies with direct measurements of many-body excitation spectra in
  an optical lattice clock.  Our work makes the first step in enabling
  a controlled, bottom-up experimental investigation of multi-body
  SU($N$) physics.
\end{abstract}

@@

\tableofcontents

@@

In this work, we investigate the first experimental capabilities with
ultracold fermionic AEAs to prepare high-density samples in a 3-D
lattice with multiple occupation of individual lattice sites
\cite{goban2018emergence}.  Specifically, we consider ground-state
preparation of isolated few-body systems in the deep-lattice limit,
and carry out a bottom-up investigation of emergent multi-body
interactions on multiply-occupied lattice sites.  These multi-body
interactions appear in a low-energy effective theory of the atoms, and
inherit the SU($N$) symmetry of their bare, pair-wise interactions,
thereby enabling experimental studies of multi-body SU($N$) physics
through the exquisite capabilities with OLCs.  Our theory is limited
to the experimental regime of at most one atom occupying each nuclear
spin state on any lattice site, which is a consequence of the
experimental protocol which starts with all atoms in the ground state.

Though effective multi-body interactions have previously been studied
in the context of harmonically \cite{johnson2012effective,
  yin2014universal} and lattice-confined \cite{johnson2009effective}
neutral bosons prepared in a single hyperfine state, our work deals
for the first time with fermions that have internal degrees of freedom
and multiple collisional parameters.  Some past work has detected
experimental signatures of multi-body interactions in the form of
quantum phase revivals \cite{will2010timeresolved}.  We instead
compare the many-body interaction energies predicted by our low-energy
effective theory to the experimental measurements of the
density-dependent orbital excitation spectra performed in
ref.~\cite{goban2018emergence}, similarly to the measurements with
bosons performed in ref.~\cite{mark2011precision}.  To facilitate this
comparison of excitation spectra and to characterize the low-lying
excitations in our effective theory, we consider a restriction of our
theory to states with at most one orbital excitation per lattice site.
In this case, we find that the SU($N$) symmetry of atomic collisions
allow the effective multi-body interactions to take a remarkably
simple form.

@@

The work in this paper is closely tied to the experimental work
reported in ref.~\cite{goban2018emergence}; we begin with a short
summary of the relevant experimental procedures therein.  The
experiment begins by preparing a degenerate gas of $10^4$-$10^5$
(fermionic) ${}^{87}$Sr atoms in a uniform mixture of their ten
nuclear spin states and at $\sim0.1$ of their fermi temperature
($\sim10$ nanokelvin) \cite{campbell2017fermidegenerate,
  marti2018imaging}.  This gas is loaded into a primitive cubic
optical lattice at the ``magic wavelength'' for which both ground
(${}^1S_0$) and first-excited (${}^3P_0$) electronic orbital states of
the atoms experience the same lattice potential \cite{ye2008quantum}.
Lattice depths along the principal axes of the lattice are roughly
equal in magnitude, with a geometric mean that can be varied from 30
to 80 $E_{\t{R}}$, where $E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is
the lattice photon recoil energy of the atoms (with the reduced Planck
constant $\hbar=1$ throughout this paper).  These lattice depths are
sufficiently large as to neglect tunneling on the time scales relevant
to the experiment.  The temperature of the atoms is also low enough to
neglect thermal occupation of motional states outside the ground-state
manifold.

Once loaded into an optical lattice, atoms are addressed by an
external (``clock'') interrogation laser with an ultranarrow (26 mHz)
linewidth, detuned by $\Delta$ from the single-atom ${}^1S_0-{}^3P_0$
transition frequency $\omega_0$.  After a fixed interrogation time,
the experiment turns off the interrogation laser, removes all
ground-state (${}^1S_0$) atoms from the lattice, and uses absorption
imaging to count the remaining excited-state (${}^3P_0$) atoms.
Non-interacting atoms in singly-occupied lattice sites feature the
typical single particle lineshape peaked at $\Delta=0$.  The
lineshapes of multiply-occupied lattice sites, meanwhile, are shifted
by inter-atomic interactions, which results in spectroscopic peaks
(i.e.~local maxima in excited-state atom counts) away from $\Delta=0$.
A sweep across different detunings $\Delta$ (on the scale of
inter-atomic interaction energies) thus constitutes a measurement of
the many-body orbital excitation spectrum.  We note that this
spectroscopic protocol addresses only singly-excited orbital states of
lattice sites.  Doubly-excited states are manifestly off-resonant for
single-photon absorption processes as $\abs{\Delta}\ll\omega_0$, while
multi-photon absorption processes are always off-resonant due to
\begin{enumerate*}
\item the interaction-induced non-linearity ($\sim$kHz) of the orbital
  excitation energies, and
\item the ultranarrow linewidth ($\sim$mHz) of the interrogation
  laser.
\end{enumerate*}

While an external trapping potential will generally break discrete
translational symmetry of the lattice, any background inhomogeneity
can be made negligible by spectroscopically addressing a sufficiently
small region of the lattice \cite{goban2018emergence}.  Throughout
this paper, we work strictly in the deep-lattice regime with
negligible tunneling between lattice sites.  We also neglect any
lattice inhomogeneities and assume that both atomic orbital states
(i.e.~${}^1S_0$ and ${}^3P_0$) experience identical lattice
potentials.  The single-particle Hamiltonian of the atoms can then be
written in the form
\begin{align}
  H_0 = \sum_{i,n,\mu,s} E_n \c_{in\mu s}^\dag \c_{in\mu s},
  \label{eq:H_0}
\end{align}
where $\c_{in\mu s}$ is a fermionic operator which annihilates a
single atom on lattice site $i\in\mathbb{Z}^3$ in motional state
$n\in\mathbb{N}_0^3$ with nuclear spin $\mu\in\set{-I,-I+1,\cdots,I}$
(i.e.~projected onto a quantization axis) and orbital state
$s\in\set{\g,\e}$; and $E_n$ is the energy of a single atom in
motional state $n$.  In a harmonic trap approximation we would have
$E_n=\p{3/2+n_\x+n_\y+n_\z}\omega$ for an on-site angular trap
frequency $\omega$, but in general the aharmonicity of the lattice
potential will cause a non-negligible shift in motional state
energies.

@@

Second, \eqref{eq:H_int_fields} is only the first term in a low-energy
expansion of two-body interactions in effective field theory, which
generally includes additional terms containing derivatives of field
operators.  Derivative terms correspond to the dependence of two-body
scattering on the relative momentum $k$ of particles involved, with
$k\to0$ in the zero-energy limit.  In the present case of $s$-wave
scattering, the leading dependence of the two-body interaction
Hamiltonian on the relative momentum $k$ can be captured by use of an
energy-dependent pseudo-potential, which amounts to using a
$k$-dependent effective scattering length \cite{blume2002fermi}.  This
effective scattering length can be determined by expanding the
$s$-wave collisional phase shift in powers of the relative momentum
$k$ \cite{giorgini2008theory, flambaum1999analytical}.  Details of
this expansion will depend on the characteristic length scale of
finite-range interactions.  In our work, these corrections to
\eqref{eq:H_int_fields} will be relevant only for the calculation of
two-body interaction energies, appearing at third order in the
coupling constants $G_X$.  As we are primarily interested in $M$-body
interactions for $M\ge3$, we defer this calculation to Appendix
\ref{sec:momentum_dependence}.  We note that our approach of using an
unregularized contact potential, renormalizing coupling constants, and
separately accounting for momentum-dependent scattering is essentially
the same as the approach used for similar calculations in
refs.~\cite{johnson2012effective, yin2014universal,
  johnson2009effective}.  While this approach does not provide insight
into the microscopic structure of inter-atomic interactions, it is
suitable for the phenomenological description of these interactions,
and particular for our eventual development of a low-energy effective
theory.

@@

becomes sharply peaked at $\ell\approx m$ as $\ell$ and $m$ get large.
The asymptotic behavior of $\alpha_3^{(3)}$ at large $\ell,m,n$ is
therefore
